Associahedra for finite type cluster algebras \\ and minimal relations between $\b{g}$-vectors

### 2019-06-13 thirdPaper/draft/typeConeAssociahedra.tex
HT comments (some of them)

This paper focuses on a surprising construction of the associahedron that recently appeared in mathematical physics.
Motivated by the prediction of the behavior of scattering particules, N.~Arkani-Hamed, Y.~Bai, S.~He, and G.~Yan recently described in~\cite[Sect.~3.2]{ArkaniHamedBaiHeYan} the kinetic associahedron.
It is a class of polytopal realizations of the classical associahedron obtained as sections of a high-dimensional positive orthant with well-chosen affine subspaces.
This construction provides a large degree of freedom in the choice of the parameters defining these affine subspaces, and actually produces all polytopes whose normal fan is affinely equivalent to that of J.-L.~Loday's associahedron~\cite{Loday} (see \cref{subsec:typeConeAsso}).
These realizations were then extended by V.~Bazier-Matte, G.~Douville, K.~Mousavand, H.~Thomas and E.~Y\i ld\i r\i m~\cite{BazierMatteDouvilleMousavandThomasYildirim} in the context of finite type cluster algebras using tools from representation theory of quivers.
More precisely, they fix a finite type cluster algebra~$\cal{A}$ and consider the real euclidean space~$\R^\cal{V}$ indexed by the set~$\cal{V}$ of cluster variables of~$\cal{A}$.
Starting from an acyclic initial seed~$\seed_\circ$, they consider a set~$\cal{M}$ of mesh mutations and compute the corresponding linear dependences between the $\b{g}$-vectors, which can be interpreted as linear spaces in~$\R^\cal{V}$.
Finally, they perturb these mesh linear spaces by a collection~$\b{\ell} \in \R^{\cal{M}}$ of positive parameters and intersect the resulting perturbed mesh affine spaces with the positive orthant of~$\R^\cal{V}$ (see \cref{subsec:typeConeCA}).
They show that the resulting polytope is always a generalized associahedron, and that its normal fan is the $\b{g}$-vector fan of~$\cal{A}$ with respect to the initial acyclic seed~$\seed_\circ$.
An implicit by-product of their construction is that the space of all polytopal realizations of the $\b{g}$-vector fan of~$\cal{A}$ with respect to the initial acyclic seed~$\seed_\circ$ is a simplicial cone.

@@

\smallskip
\noindent
These two aspects are interesting in their own right and might be useful in different communities: the geometric side enables to simply describe all polytopal realizations of a fan with simplicial type cone (which is sometimes easier to check than to find an explicit realization), while the algebraic side provides fundamental new results on the $\b{g}$-vectors of cluster algebras and on (an extriangulated variant of) the Grothendieck groups of~$2$-Calabi--Yau triangulated categories with cluster tilting objects.
We have therefore voluntarily split our presentation into two clearly marked parts that can be read independently.
The geometric side, which uses elementary techniques, is presented in \cref{part:geometry} together with all its consequences on polytopal realizations.
The algebraic side, which is of independent interest, is presented in \cref{part:algebra}.

@@

\medskip
To conclude, it is our hope that the present paper will participate to the interactions between combinatorial geometry, representation theory, and mathematical physics; in particular through further developments of the amplituhedron program, which was open in~\cite{ArkaniHamedTrnka-Amplituhedron} and has flourished during the last years~\cite{ArkaniHamedBourjailyCachazoGoncharovPostnikovTrnka-GrassmannianGeometryScatteringAmplitudes}. The kinetic associahedron from~\cite{ArkaniHamedBaiHeYan} is one of the last ``positive geometries''~\cite{ArkaniHamedBaiLam-PositiveGeometries} that has emerged in the study of scattering amplitudes.

>>>>

This paper focuses on a surprising construction of the associahedron that recently appeared in mathematical physics.
Motivated by the prediction of the behavior of scattering particles, N.~Arkani-Hamed, Y.~Bai, S.~He, and G.~Yan recently described in~\cite[Sect.~3.2]{ArkaniHamedBaiHeYan} the kinematic associahedron.
It is a class of polytopal realizations of the classical associahedron obtained as sections of a high-dimensional positive orthant with well-chosen affine subspaces.
This construction provides a large degree of freedom in the choice of the parameters defining these affine subspaces, and actually produces all polytopes whose normal fan is affinely equivalent to that of J.-L.~Loday's associahedron~\cite{Loday} (see \cref{subsec:typeConeAsso}).
These realizations were then extended by V.~Bazier-Matte, G.~Douville, K.~Mousavand, H.~Thomas and E.~Y\i ld\i r\i m~\cite{BazierMatteDouvilleMousavandThomasYildirim} in the context of finite type cluster algebras using tools from representation theory of quivers.
More precisely, they fix a finite type cluster algebra~$\cal{A}$ and consider the real euclidean space~$\R^\cal{V}$ indexed by the set~$\cal{V}$ of cluster variables of~$\cal{A}$.
Starting from an acyclic initial seed~$\seed_\circ$, they consider a set~$\cal{M}$ of mesh mutations and compute the corresponding linear dependences between the $\b{g}$-vectors, which can be interpreted as linear spaces in~$\R^\cal{V}$.
Finally, they perturb these mesh linear spaces by a collection~$\b{\ell} \in \R^{\cal{M}}$ of positive parameters and intersect the resulting perturbed mesh affine spaces with the positive orthant of~$\R^\cal{V}$ (see \cref{subsec:typeConeCA}).
They show that the resulting polytope is always a generalized associahedron, and that its normal fan is the $\b{g}$-vector fan of~$\cal{A}$ with respect to the initial acyclic seed~$\seed_\circ$.
An implicit by-product of their construction is that the space of all polytopal realizations of the $\b{g}$-vector fan of~$\cal{A}$ with respect to the initial acyclic seed~$\seed_\circ$ is a simplicial cone.

@@

\smallskip
\noindent
These two aspects are interesting in their own right and might be useful in different communities: the geometric side enables to simply describe all polytopal realizations of a fan with simplicial type cone (which is sometimes easier to check than to find an explicit realization), while the algebraic side provides fundamental new results on the $\b{g}$-vectors of cluster algebras and on (an extriangulated variant of) the Grothendieck groups of~$2$-Calabi--Yau triangulated categories with cluster tilting objects.
We have therefore deliberately split our presentation into two clearly marked parts that can be read independently.
The geometric side, which uses elementary techniques, is presented in \cref{part:geometry} together with all its consequences on polytopal realizations.
The algebraic side, which is of independent interest, is presented in \cref{part:algebra}.

@@

\medskip
To conclude, it is our hope that the present paper will participate to the interactions between combinatorial geometry, representation theory, and mathematical physics; in particular through further developments of the geometric approach to scattering amplitudes that has flourished during the last years~\cite{ArkaniHamedTrnka-Amplituhedron,ArkaniHamedBourjailyCachazoGoncharovPostnikovTrnka-GrassmannianGeometryScatteringAmplitudes}. The kinematic associahedron from~\cite{ArkaniHamedBaiHeYan} is one of the most recent ``positive geometries''~\cite{ArkaniHamedBaiLam-PositiveGeometries} that has emerged in the study of scattering amplitudes.
